Data Specialization for a Staged Lambda Calculus

### 2014-11-07 paper/splitting.tex
Edits through 4.3

\subsection{Splitting Stage \bbone\ Terms}
\label{sec:split-one}

@@

The splitting algorithm for stage~\bbone\ terms, as specified by the
judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplitOne}, proceeds
recursively on the structure of~$e$.
When $e$ is a terminal (a variable or base constant)
splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}. (Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.

@@

\subsection{Splitting Stage \bbtwo\ Terms}

Rather than match the

Splitting stage~\bbtwo\ terms in \lang\ takes on a simpler form than stage~\bbone\ terms since stage~\bbtwo\ terms because rather that the output of splitting

Like stage one, we get two terms.

But now the objective is to match the partial value, but the value after full evaluation.

That means we need to match values.

Splitting a stage~\bbtwo\ term $e$ in \lang\ generates a
precomputation~$p$ that performs all stage~\bbone\ subcomputations in $e$,
and a~resumer function $e_2=l.r$ that performs all stage~\bbtwo\ subcomputations.
If full evaluation of $e$ in~\lang\ reduces to the value~$v$, then $\letin l p r$ must reduce to~$v$ in \langmono.

@@

As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms
as specified by the
judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, proceeds
recursively on the structure of~$e$.  In the terminal cases of
constants and variables, splitting generates \texttt{()} precomputations (they do not involve stage~\bbone\
subcomputations), and resumers that produce the original term. For
example, the integer constant \texttt{5} splits into the
precomputation \texttt{()} and \texttt{l=>5}.

For all non-terminal stage~\bbtwo\ terms~$e$ in \lang\ (except $\prev$
and $\pause$), splitting descends into~$e$, recursively splitting
subterms and generating precomputations $p_1, \ldots, p_n$ and
resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
The precompuation of~$e$ is defined as the combined precomputations:
$p=(p_1,\ldots,p_n)$.  The resumer binds each precomputation to an
argument $(l_1,\ldots,l_n)$ in a term that has the same structure
of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).
Notice that splitting \texttt{case}s and functions causes the
precomputation of subterms to be lifted out from underneath stage
\bbtwo\
binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
\ur{This last point seems important.  It would be important to come
  back to it.}

\subsection{Interface Rules}

The rule for $\next$ simply tuples up the precomputation of its subterm with a trivial immediate result,
while the rule for $\prev$ projects the combined result of its subterm to isolate the precomputation.
Since the argument to $\prev$ must have a $\fut$ type, its immediate result must reduce to \texttt{()},
justifying why that part can be thrown away.
The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
and projects out the integer result in the resumer.% 
\footnote{The resumer of an integer expression is usually trivial, 
but we have to include it here for termination purposes.}

%For example, the stage \bbtwo\ term
%\begin{lstlisting}
%2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
%\end{lstlisting}
%has two subterms, the latter of which itself has two subterms. Two are $\pause$s
%containing stage \bbone\ computations, while the third has no stage \bbone\
%content. 
%
%\ur{The discussion of the subterms is not helpful. what is the point?}
%
%The precomputation \emph{lifts} these computations in the same tree
%structure: \verb|(1+2,(3+4,()))|.%
%\ur{Not clear what a ``tree structure'' is. }
%\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
%The resumer is a function which computes the stage \bbtwo\ value of the original
%term by mapping the results of the precomputation in place of the $\pause$s:
%\verb|(L1,(L2,_)) => L1 < (L2 - 5)|. 
%
%\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
%
%Non-staging constructs with $n$ subterms have precomputations which are
%$n$-tuples of their subterms' precomputations, and resumers which pass each
%component to the resumer of the respective subterm.

@@

\subsection{Optimizations}

@@

\subsection{Implementation}

\crem{Say something about the implementation.}

>>>>

\subsection{Stage \bbone\ Term Splitting}
\label{sec:split-one}

@@

The splitting algorithm for stage~\bbone\ terms, as specified by the
judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplitOne}, proceeds
recursively on the structure of~$e$.
When $e$ is a terminal (a variable or base constant)
splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}. (Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.

@@

Since the results of splitting $\next$ terms depend on the output of splitting its stage~\bbtwo\ subterm,
we defer description of $\next$ until after describing stage~\bbtwo\ term splitting. % in the subsequent section.

\subsection{Stage \bbtwo\ Term Splitting}

Because stage~\bbtwo\ terms in \lang\ fully reduce to values (as opposed to partial values),
splitting stage~\bbtwo\ terms in \lang\ assumes a simpler form than that of stage~\bbone\ term splitting. 
Specifically, splitting a stage~\bbtwo\ term $e$ in \lang\ generates a precomputation term $p$
(performing all stage~\bbone\ subcomputations in $e$) and a
resumer function $e_2=l.r$ (performing all stage~\bbtwo\ subcomputations) such that when 
$e$ reduces to the value~$v$, $\letin l p r$ reduces to an identical value in \langmono.
Note that in contrast to the combined term produced during stage~\bbone\ splitting,
the precomputation $p$ generated by stage~\bbtwo\ splitting exists entirely
to compute the input expected by the resumer.

@@

As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms,
as specified by the judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, is defined on the local structure of $e$.
In the terminal cases of
constants and variables, splitting generates trivial precomputations that are \texttt{()}, and resumers that produce the original term. For
example, the integer constant \texttt{3} splits into the
precomputation \texttt{()} and resumer \texttt{\_=>3}.

For non-terminal terms~$e$ in \lang\ (except $\prev$
and $\pause$)
% recursively splitting
% subterms and generating precomputations $p_1, \ldots, p_n$ and
% resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
the precomputation of~$e$ is defined as the combined precomputations of $e$'s $n$ subterms:
$p=(p_1,\ldots,p_n)$.  The resumer binds each precomputation to an
argument $(l_1,\ldots,l_n)$ in a term that has the same structure
of~$e$ but where each subterm is replaced by its corresponding resumer $r_i$.
A notable properly of splitting stage~\bbtwo\ \texttt{case}s and functions is that the
precomputation of subterms is lifted out from underneath stage
\bbtwo\
binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
\ur{This last point seems important.  It would be important to come
  back to it.}

Splitting $\prev$ generates a precomputation that projects the combined result of its stage~\bbone\ subterm.
Since the argument to $\prev$ is of $\fut$ type, its immediate result reduces to \texttt{()}, justifying why it can be thrown away.
$\pause$ treats the entire combined result of its stage~\bbone\ subexpression as a precomputation, 
and projects out the integer result in the resumer. 
\footnote{The resumer of an integer expression is usually trivial, 
but we have to include it here for termination purposes.} 
Finally, splitting $\next$ simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial immediate result \texttt{()}.

@@

\subsection{Optimization: Monostage Functions}

\textbf{KF: the missing point here is that the splitting algorithm was designed based on the local structure of \lang\ terms.
To make the algorithm derivation case compelling,
we need to optimize away some of that structure.  But this is hard to do through functions.}

@@

% \subsection{Implementation}
%
% \crem{Say something about the implementation.}
